Tail Flexibility in the Degrees of Preferential Attachment Networks

### 2025-05-21 paper.tex
Updating draft

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Tail Flexibility in the Degrees of Preferential Attachment Networks},
  pdfauthor={Thomas William Boughen; Clement Lee; Vianey Palacios Ramirez},
  pdfkeywords={networks, extremes},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\section{Introduction}\label{sec-intro}

Networks appear in many fields such as sociology, politics,
epidemiology, and economics. Statistical methods have been and continue
to be used to study networks that appear in these fields and beyond,
from the use of stochastic block models to detect communities within the
French political blogosphere \citep{Latouche11}, to using Exponential
Random Graph Models (ERGMs) to analyse the global trade network
\citep{Setayesh22}, and the use of mechanistic models in neuroscience to
investigate the patterns of connections in neural systems
\citep{Betzel17}. One such mechanistic model is the general preferential
attachment (PA) model where as a network grows and new vertices join
they connect to those already present in the network with weights
proportional to some preference function \(b\) of their degree. The
well-known Barabási-Albert (BA) model is a special case of this when
\(b\sim k\) {[}BA99{]}.

Mechanistic models, like the PA model, are in general only particularly
useful when simulating networks and not fitting to the real data as
usually only snapshots of real networks are available and not the entire
evolution. As such, one of the most informative statistics when
regarding a network is the degree distribution. When looking at the
degree distributions of real networks it is attractive to many to use
the power law to model the degree distribution, as the BA model
generates degrees that follow a power law, and therefore claim that
since the degree distribution is roughly power law then the underlying
growth mechanism may be preferential attachment such as the BA model.
However, it has been shown that a general preference function will not
always lead to a power law degree distribution \citep{krapivsky01} and
additionally it has been debated for decades whether most real data
actually follows a power law \citep{Voitalov_2019} or
not\citep{Broido_2019}.

Throughout this debate the use of methods from extremes has been absent
until relatively recently with the papers such as \citep{Voitalov_2019}
and {[}Lee{]}. Forgoing the use of extreme value methods essentially
downplays how deviations from the power law in the largest degrees may
affect the network given that they are often much more influential than
the vertices with smaller degrees.

Noting this, many have now started to use these extreme value methods
when analysing networks. \citep{wan19} showed that extreme value
estimation methods prove more robust than existing parametric approaches
especially when the data are corrupted or the model has been
misspecified, demonstrating that extreme value methods can be a valuable
tool when studying networks. \citep{wang2022random} investigates a PA
model with a affine preference function (similar to the BA model) with
the added behaviour of heterogeneous reciprocal edges, they show that
the distribution of in-degrees and out-degrees produced by such a model
is multivariate regularly varying and they also have hidden regular
variation. {[}Lee{]} using methods from discrete extremes modelled the
right tail of network degrees with a discretised variation of the
generalised Pareto distribution (GPD) usually dubbed the Integer GPD
(IGPD) where \(X|X>v \sim \mathrm{IGPD}(\xi,\sigma,v)\) is defined by
the survival:

\[
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
\] for \(v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R\).

The study of degree distributions, whether using extremes or not,
usually does not reveal any information about the preference function
even if PA was the underlying generating mechanism of the network. Here
we address this gap in the literature, more specifically investigating
if given the degree distribution of a network assumed to come from the
PA model can we directly infer the model parameters? We build upon the
foundation provided by \citep{rudas07} deriving the limiting degree
distribution in terms of the preference function and suggest a class of
preference functions that result in realistic tails that are heavy but
are lighter than implied by the power law.

\subsection{Structure of this paper}\label{structure-of-this-paper}

\textbf{?@sec-gpa} looks at the theoretical limiting behaviour of a GPA
model when \(m=1\) according to the results from \citep{rudas07}, going
on to introduce a flexible class of preference functions that can
guarantee a heavy tailed degree distribution while still being flexible
in the body. Section~\ref{sec-rec} shows that the preference function
parameters can be fairly well recovered from the degree distributions of
networks simulated from the GPA model using various functions of the
class introduced in \textbf{?@sec-gpa}. Having shown that for simulated
data the model parameters are recoverable from the degree distribution
alone, Section~\ref{sec-real} builds upon this and attempts to fit the
model to degree distributions of real networks; modelling not only the
degree distribution itself but also the estimates for the parameters of
the preference function assuming the networks evolved according to the
GPA scheme. Ultimately, Section~\ref{sec-conc} discusses the main
results of the paper, pitfalls of this method and future work.

\section{PA model}\label{pa-model}

@@

\[
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)},\qquad k\in\mathbb N.
\] We are specifically interested in how the tail heaviness (\(\xi\) in
the IGPD) is affected by the preference function \(b\),
\citep{shimura12} introduces a quantity that will help in determining
what domain of attraction this discrete distribution belongs to, that
is:

For a distribution \(F\) with survival function \(\bar F\) and some
\(n\in\mathbb Z^+\) let:

\[
\Omega(F,n) = \left(\log\displaystyle\frac{\bar F (n+1)}{\bar F (n+2)}\right)^{-1} - \left(\log\displaystyle\frac{\bar F (n)}{\bar F (n+1)}\right)^{-1}
\]

\citep{shimura12} then states that if
\(\lim_{n\rightarrow\infty} \Omega(F,n) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is heavy tailed with \(\bar F(n) \sim n^{-\alpha}\).
Additionally, if \(\lim_{n\rightarrow\infty} \Omega(F,n) = 0\) then the
distribution is light tailed. This allows us to show the following

@@

See \hyperref[appendix]{appendix} for the details of the proof.

@@

We can show that in order for the degree distribution to be heavy
tailed, the preference function must be asymptotically linear
i.e.~\(\lim_{k\rightarrow\infty}\frac{b(k)}{k} =  c>0\).

Consider the definition of heavy tails:

\begin{align*}
\lim_{k\rightarrow\infty}\Omega(F,n) &= c>0\\
\lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*} &= c\\
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] &= \lambda^*c\\
\lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{(k+1)-(k)} &= \lambda^*c\\
\end{align*}

by Stolz-Cesaro Theorem:

\[
\lim_{k\rightarrow\infty}\frac{b(k)}{k} = \lambda^*c > 0.
\]

Since \(b\) must be asymptotically linear in order to have positive tail
heaviness we propose the following preference function:

@@

\section{Preferential Attachment with flexible heavy
tail}\label{sec-model}

@@

\[
\hat\rho(\lambda) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda + i^\alpha + \varepsilon} 
\]

which must be solved numerically for most parameter choices. Also, note
that \(\lambda>\beta\) since when \(\lambda\le \beta\) the infinite sum
no longer converges and instead goes to infinity.

@@

@@

The survival (Equation~\ref{eq-polysurv}) can be connected to the IGPD
mentioned in Section~\ref{sec-intro} by using Stirling's approximation
of the gamma function to obtain:

\[
\frac{\Gamma(x+y)}{\Gamma(x)}\approx x^{y},
\] and utilising the expression from Equation~\ref{eq-polysurv}:

@@

\section{Recovery}\label{sec-rec}

The goal of was to provide a possible preference function that is able
to explain the growth of real networks with varied tail behaviour in
their degree distributions. This section aims to show that the
parameters of the model in Section~\ref{sec-model} can be recovered from
the degree distribution of a network simulated from it.

The procedure for recovering the parameters begins with simulating a
network from the model with \(N=100,000\) vertices and \(m=1\) given
some set of parameters
\(\pmb\theta = (\alpha, \beta, \varepsilon, k_0)\), obtaining the degree
counts in the form of a vector of degrees \(\pmb x = (1,2,\ldots,M)\)
and the number of vertices with those degrees \(\pmb n\) where \(M\) is
the maximum degree. However, when it comes to fitting this model to real
data in the next section, small degrees are extremely influential and
will often cause the model to have a bad fit even though the rest of the
data looks similar to that of the simulations; truncating the data above
some value, say \(l<k_0\), should allow the model to be more effectively
fitted. This accounts in some way for the fact that this model only
applies to trees, and could not possibly capture the behaviour for nodes
with small degrees. Using the p.m.f derived in \citep{rudas07}, the
likelihood given that \(x_i \ge l\) for all \(i =1,2,\ldots,M\) is:

\begin{align*}
L(\pmb x,\pmb n | \pmb \theta) = \left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{x_i\ge k_0}n_{x_i}\right)} \prod_{l \le x_i<k_0}\left(\frac{\lambda^*}{\lambda^* +x_i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ \times \prod_{x_i\ge k_0}\left(\frac{\text{B}(x_i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{align*}

where \(B(y,z)\) is the the beta function.

This likelihood allows inference to be made about the parameters using a
Bayesian approach using the priors:

@@

\captionsetup{labelsep=none}

{\centering \includegraphics{mcmc_plot.jpg}

@@

\caption{: Posterior estimates of paramters for data simulated from the
proposed model with various combinations of
(\(\alpha\),\(\beta\),\(\varepsilon\)) and \(k_0=20\).}

\caption{\label{fig-rec2}}

@@

Turning now to real data, the goal is to fit the model to the degree
distributions of real networks from various sources. {[}describe
networks being used{]}. Alongside fitting this model to the degree
distributions, we then compare the fit to that of an mixture
distribution that was used in {[}clement{]} and is similar to others
that have been used to model degree distributions {[}others{]}.

@@

\captionsetup{labelsep=none}

{\centering \includegraphics{pars_plot.png}

@@

\caption{\label{fig-rec2}}

\section{Conclusion and Discussion}\label{sec-conc}

@@

GPA is also a fairly simplistic one on its own despite having large
flexibility in the preference function allowing for edges to only be
added at times when a new node is introduced into the network and not
allowing for anything like vertex/edge death, reciprocal edges and
rewiring steps. \citep{deijfen2015} introduces some theory for GPA with
vertex death based on the work by \citep{rudas07}, but is unable to come
to an expression for the degree distribution. However, adding something
like fitness scores into the GPA model seems fairly simple to do even
going so far as making each of the parameters used in the class of
functions here node-wise instead of universal. For example having the
preference weight for a vertex be given by
\(b(k_i, \underline{\gamma_i})\) where \(\underline{\gamma_i}\sim G\) is
the vertex's fitness and \(k_i\) is the vertex's degree, it is
reasonable to expect that the degree distribution for a model like this
to be given by:

\[
\bar F(k) = \mathbb E_G\left[\prod_{i=0}^k \frac{b(i, \gamma)}{\lambda^* + b(i,\gamma)}\right],\qquad \gamma\sim G,\, k=0,1,2,\ldots
\]

If this were the case, the posterior distributions of parameters from
Section~\ref{sec-real} could be interpreted as the approximate
distributions from which a vertex's fitness values are drawn instead of
providing estimates of the universal parameter values. Additionally, it
allows the ability to set some parameters as universal and some as
vertex-wise for example one could let \(\alpha, \beta, k_0\) be the same
for all vertices but let each vertex in the network have its own ``zero
appeal''
\(\varepsilon_i\overset{\text{iid}}{\sim} \text{Ga}(\theta_1, \theta_2)\);
moving the model to being more realistic and accounting for the fact
that in many real networks the objects vertices represent have some
value instrinsic to themselves other than the number of connections they
have.

\section{Appendix}\label{appendix}

\subsection{Additional Results}\label{additional-results}

\subsubsection{Tail heaviness of general preferential attachment
model}\label{tail-heaviness-of-general-preferential-attachment-model}

Recall the limiting survival function:

\[
\bar F(k) = \prod_{i=0}^k \frac{b(i)}{\lambda^*+b(i)},
\]

The aim is to determine how this distribution behaves for different
choices of \(b\), specifically what maximum domain of attraction does
this distribution belong to and is it affected by the choice of
preference function \(b\). \citet{shimura12} introduces a quantity that
will help in determining what domain of attraction a discrete
distribution belongs to, that is:

For a distribution \(F\) with survival function \(\bar F\) and some
\(n\in\mathbb Z^+\) let:

\[
\Omega(F,n) = \left(\log\displaystyle\frac{\bar F (n+1)}{\bar F (n+2)}\right)^{-1} - \left(\log\displaystyle\frac{\bar F (n)}{\bar F (n+1)}\right)^{-1}
\]

\citep{shimura12} then states that if
\(\lim_{n\rightarrow\infty} \Omega(F,n) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is heavy tailed with \(\bar F(n) \sim n^{-\alpha}\).
Additionally, if \(\lim_{n\rightarrow\infty} \Omega(F,n) = 0\) then the
distribution is light tailed.

Consider a preference function \(b(\cdot)\) that fulfills:

\begin{equation}\phantomsection\label{eq-condb1}{
\lim_{n\rightarrow \infty} b(n) = \infty
}\end{equation}

Substituting in the form of \(\bar F(n)\) from \textbf{?@eq-surv} and
taking the limit, subject to Equation~\ref{eq-condb1}:

\begin{equation}\phantomsection\label{eq-omegares}{
\lim_{n\rightarrow\infty}\Omega(F,n) = \lim_{n\rightarrow\infty}\frac{b(n+2)-b(n+1)}{\lambda^*}.
}\end{equation} (see appendix for full proof).

Whilst there are many classes of functions that satisfy
Equation~\ref{eq-condb1} and Equation~\ref{eq-condb2}, the vast majority
of these result in \(\Omega(F,n) \rightarrow 0\), meaning that the
limiting degree distribution is light tailed. In fact, in order for the
limiting degree distribution to be heavy tailed, \(b\) must be
asymptotically linear (\(b(k)\sim k\)).

\subsection{Proofs}\label{proofs}

\subsubsection{Tail heaviness of GPA}\label{tail-heaviness-of-gpa}

@@

\begin{align*}
\Omega(F,n) &= \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}+\frac{b(n+2)}{\lambda}-\left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}+\frac{b(n+1)}{\lambda}  -\frac{b(n+1)}{\lambda}\\
&=\left\{ \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}\right\} - \left\{ \left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}-\frac{b(n+1)}{\lambda}\right\}+\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}
\end{align*}

@@

\begin{align*}
\lim_{n\rightarrow\infty}\Omega(F,n) &= \lim_{n\rightarrow\infty}\left\{ \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}\right\} - \lim_{n\rightarrow\infty}\left\{ \left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}-\frac{b(n+1)}{\lambda}\right\}\\
&\qquad+\lim_{n\rightarrow\infty}\left(\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}\right)\\
&=\frac{1}{2}-\frac{1}{2} + \lim_{n\rightarrow\infty}\left(\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}\right)\\
&=\frac{1}{\lambda}\lim_{n\rightarrow\infty}\left[b(n+2)-b(n+1)\right]\qquad \square
\end{align*}

\subsubsection{Removing the infinite
sum}\label{removing-the-infinite-sum}

@@

\section{References}\label{references}

>>>>

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Tail Flexibility in the Degrees of Preferential Attachment Networks},
  pdfauthor={Thomas William Boughen; Clement Lee; Vianey Palacios Ramirez},
  pdfkeywords={networks, extremes},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\newpage

\section{Introduction}\label{introduction}

Networks appear in many fields such as sociology, politics,
epidemiology, and economics. Statistical methods have been and continue
to be used to study networks that appear in these fields and beyond,
from the use of stochastic block models to detect communities within the
French political blogosphere \citep{Latouche11}, to using Exponential
Random Graph Models (ERGMs) to analyse the global trade network
\citep{Setayesh22}, and the use of mechanistic models in neuroscience to
investigate the patterns of connections in neural systems
\citep{Betzel17}.

One example of a mechanistic model that may be considered is the
preferential attachment (PA) model where as a network grows and new
vertices join they connect to those already present in the network with
weights proportional to some preference function \(b(\cdot)\) of their
in-degree. The well-known Barabási-Albert (BA) model is a special case
of this when \(b(k) = k+1\) \citep{Barabasi99}.

These mechanistic models are generally useful when simulating networks,
but not for fitting to real data as usually only snapshots of a network
are available and not the full evolution. However, one statistic that is
usually available for snapshots of networks is the degree distribution
making it valuable to study the degrees in a network. Often, when
looking at the degree distributions of real networks it can be
attractive to use the power law when modelling as it is seemingly
observed in much of the data. It also comes with the added benefit of
offering a connection to preferential attachment since the BA model
results in a power law degree distribution \citep{Barabasi99}, perhaps
suggesting that if the power law fits well then the network may have
grown according to similar simple rules. However, attempting to explain
the networks growth using preferential attachment when the degrees
seemingly follow a power law may not be the best course of action as
\citep{krapivsky01} shows that a general preference function does not
lead to power law degrees and additionally, it is still being argued
whether or not a lot of real data actually follows a power law.
\citet{Broido_2019} provides some evidence that suggests that real world
networks rarely have degrees that follow a power law through the use of
statistical analysis of almost 1000 networks across a wide variety of
fields; \citet{Voitalov_2019}, on the other hand, disagrees and claims
that these kinds of networks are not nearly as rare and only appear so
as a result of an unrealistic expectation of power law without
deviations or noise.

Extreme value methods have been scarcely used when it comes to analysis
of degree distributions and whether they follow the power law. The
absence of these methods essentially down weights the deviation of
vertices with the largest degrees, which are usually the most
influential, from the power law. Somewhat recent publications
\citep{Voitalov_2019, Lee24, wang2022random} have addressed this gap in
the literature. \citet{Voitalov_2019} proposes estimating the power law
exponent of a degree distribution using three estimators based in
extreme value theory, in contrast to the estimators generally used in
network science, the degree distributions are then split into one of
three categories (not power-law , hardly power-law or power-law)
depending on the values of the estimates obtained from each of the
estimators. \citet{Lee24} proposes a framework for modelling entire
degree distributions using a combination of two generalisations of the
power law including a model selection step that is capable of suggesting
whether or not the data is adequately modelled by a power law, with one
conclusion being that most data seems to have a tail lighter than what
is implied by the body whilst still being heavy tailed.
\citet{wang2022random} considers a preferential attachment model with an
affine preference function and reciprocal edges, they study the limiting
degree distribution of this model for both the in-degrees and
out-degrees - finding that the joint limiting distribution is
multivariate regularly varying and has the property of hidden regular
variation.

Evidently, the study of degree distributions is a very active but the
vast majority of research in the field (whether using extremes or not)
are purely descriptive in the sense that no information about the
preference function is revealed even if preferential attachment was the
underlying mechanism. This paper aims to address this gap - asking if
given the degree distribution of a network assumed to come from the
preferential attachment model, can we directly infer the model
parameters?

\citet{rudas07} derives the limiting degree distribution in terms of the
preference function \(b(\cdot)\) and its parameters, we build upon this
by finding a class of preference functions that result in realistic
tails i.e.~heavy but not as heavy as implied by the power law.

The remainder of this paper begins with a more detailed description of
the PA model alongside the theoretical results for the limiting degree
distribution using results from \citet{rudas07}, this section
(Section~\ref{sec-tail}) then goes on to analyse the tail behaviour of
this limiting distribution in terms of the preference function
\(b(\cdot)\) arguing that a heavy tailed degree distribution can only be
the result of an asymptotically linear preference function - ending with
a suggestion for a preference function that can guarantee heavy tails
while remaining flexible. Section~\ref{sec-model} continues on from the
previous section by considering the limiting distribution obtained when
using a preference function of the form stated at the end of
Section~\ref{sec-tail} providing examples of how the shape of the
distribution varies with respect to the model parameters as well as
visualisations for how the tail heaviness changes when these parameters
are varied - continued study of this distribution then also reveals a
connection to the general Pareto distribution (GPD).
Section~\ref{sec-sim} consists of a simulation study where networks are
simulated from the PA model for various parameter combinations and then
the model specified in Section~\ref{sec-model} is fitted to the
resulting degree distributions using Bayesian inference, demonstrating
that (for simulated data) the model parameters can be recovered from
only the degree distribution. Finally, Section~\ref{sec-real} fits the
model to some real data and provides posterior estimates for the
preference function assuming the networks evolved according to the PA
model.

\newpage

\section{Tail Behaviour of Preferential Attachment
Model}\label{sec-tail}

@@

\[
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)},\qquad k\in\mathbb N.
\]

We are specifically interested in how the tail heaviness (\(\xi\) in the
IGPD) is affected by the preference function \(b\), \citep{shimura12}
introduces a quantity that will help in determining what domain of
attraction this discrete distribution belongs to, that is:

For a distribution \(F\) with survival function \(\bar F\) and some
\(k\in\mathbb Z^+\) let:

\[
\Omega(F,k) = \left(\log\displaystyle\frac{\bar F (k+1)}{\bar F (k+2)}\right)^{-1} - \left(\log\displaystyle\frac{\bar F (k)}{\bar F (k+1)}\right)^{-1}
\]

\citep{shimura12} then states that if
\(\lim_{n\rightarrow\infty} \Omega(F,k) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is heavy tailed with \(\bar F(k) \sim k^{-\alpha}\).
Additionally, if \(\lim_{n\rightarrow\infty} \Omega(F,k) = 0\) then the
distribution is light tailed. This allows us to show the following:

@@

See appendix \ref{sec-proofs} for the details of the proof.

@@

We can show that in order for the degree distribution to be heavy
tailed, the preference function must be asymptotically linear
i.e.~\(\lim_{k\rightarrow\infty}\frac{b(k)}{k} =  c>0\). First we must
consider the following theorem:

\begin{theorem}[Stolz-Cesàro Theorem
\citep{cesaro}]\protect\hypertarget{thm-stolz}{}\label{thm-stolz}

Let \((a_n)_{n\ge1}\) and \((b_n)_{n\ge1}\) be two sequence of real
numbers. Assume \((a_n)_{n\ge1}\) is a strictly monotone and divergent
sequence and the following limit exists:

\[
\lim_{k\rightarrow\infty}\frac{b_{k+1} - b_k}{a_{k+1} - a_k} = l.
\] Then,

\[
\lim_{k\rightarrow\infty}\frac{b_k}{a_k} = l.
\]

\end{theorem}

From Proposition~\ref{prp-omega} we know that in order for the limiting
degree distribution to have heavy tails, we must have that:

\[
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = c>0.
\] Now, setting \(b_k = b(k)\) and \(a_k = k\):

\[
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = \lim_{k\rightarrow\infty}\frac{b_{k+1} - b_k}{a_{k+1} - a_k} = c>0,
\] meaning by Theorem~\ref{thm-stolz}:

\[
\lim_{k\rightarrow\infty}\frac{b_k}{a_k} = \lim_{k\rightarrow\infty}\frac{b(k)}{k} = c>0.
\] Therefore, in order for the degree distribution to be heavy tailed,
the preference function must be asymptotically linear. With this under
consideration we propose the following preference function:

@@

\newpage

\section{Preferential Attachment with Flexible Heavy
Tail}\label{sec-model}

@@

\begin{equation}\phantomsection\label{eq-rho}{
\hat\rho(\lambda) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda + i^\alpha + \varepsilon} 
}\end{equation}

which must be solved numerically for most parameter choices. Also, note
that \(\lambda>\beta\).

@@

\begin{figure}

@@

The survival (Equation~\ref{eq-polysurv}) can be connected to the IGPD
by using Stirling's approximation of the gamma function, where the IGPD
is a discretised variation of the generalised Pareto distribution (GPD)
usually dubbed the Integer GPD (IGPD) where
\(X|X>v \sim \mathrm{IGPD}(\xi,\sigma,v)\) defined by the survival: \[
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
\] for \(v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R\).

Using Stirling's approximation we have that: \[
\frac{\Gamma(x+y)}{\Gamma(x)}\approx x^{y},
\] now utilising the expression from Equation~\ref{eq-polysurv}:

@@

\section{Simulation Study}\label{sec-sim}

This section aims to show that the parameters of the model in
Section~\ref{sec-model} can be recovered from the degree distribution of
a network simulated from it.

The procedure for recovering the parameters begins with simulating a
network from the model with \(N=100,000\) vertices and \(m=1\) given
some set of parameters
\(\pmb\theta = (\alpha, \beta, \varepsilon, k_0)\), obtaining the degree
counts and using the likelihood from the previous section alongisde the
priors:

@@

\pandocbounded{\includegraphics[keepaspectratio]{images/mcmc_plot.jpg}}

@@

\caption{\label{fig-rec2}: Posterior estimates of paramters for data
simulated from the proposed model with various combinations of
(\(\alpha\),\(\beta\),\(\varepsilon\)) and \(k_0=20\).}

@@

Turning now to real data, the goal is to fit the model to the degree
distributions of real networks from various sources. Alongside fitting
this model to the degree distributions, we then compare the fit to that
of an mixture distribution that was used in \citet{Lee24}.

\begin{figure}

@@

{\centering \pandocbounded{\includegraphics[keepaspectratio]{images/pars_plot.png}}

@@

\section{Conclusion and Discussion}\label{conclusion-and-discussion}

@@

\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}
\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{A\arabic{figure}}

\newpage

\section{Proofs and Derivations}\label{sec-proofs}

\subsection{\texorpdfstring{Proof of
Proposition~\ref{prp-omega}}{Proof of Proposition~}}\label{proof-of-prp-omega}

@@

\begin{align*}
\Omega(F,n) &= \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}+\frac{b(n+2)}{\lambda}-\left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}\\&+\frac{b(n+1)}{\lambda}  -\frac{b(n+1)}{\lambda}\\
&=\left\{ \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}\right\} - \left\{ \left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}-\frac{b(n+1)}{\lambda}\right\}\\&+\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}
\end{align*}

@@

\begin{align*}
\lim_{n\rightarrow\infty}\Omega(F,n) &= \lim_{n\rightarrow\infty}\left\{ \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}\right\} \\&- \lim_{n\rightarrow\infty}\left\{ \left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}-\frac{b(n+1)}{\lambda}\right\}\\
&\qquad+\lim_{n\rightarrow\infty}\left(\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}\right)\\
&=\frac{1}{2}-\frac{1}{2} + \lim_{n\rightarrow\infty}\left(\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}\right)\\
&=\frac{1}{\lambda}\lim_{n\rightarrow\infty}\left[b(n+2)-b(n+1)\right]\qquad \square
\end{align*}

\subsection{\texorpdfstring{Derivation of
Equation~\ref{eq-rho}}{Derivation of Equation~}}\label{derivation-of-eq-rho}

@@

\newpage

\section{Additional Tables and
Figures}\label{additional-tables-and-figures}

\newpage

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}
